---
title: "Crime Rates"
---

Follow along with the demo to learn how to calculate crime rates using R and the Tidyverse. We'll also introduce a new library, janitor, which helps make column names easier to work with. You should get started by loading the tidyverse in a new R block, then install and load janitor.

### Task 1: Load the tidyverse and install and load janitor

```{r}
library(tidyverse)
install.packages("janitor")
library(janitor)
```



### Task 2: Load the crime dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_annual_crime.csv

```{r}
crimes <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_annual_crime.csv") |> clean_names()
```


### Task 3: Use janitor to clean up the column names
```{r}
clean_names(crimes)
```



### Task 4: Calculate the total crime rate for each year (picking our unit!)

```{r}
crimes |> mutate(total_rate = grand_total/population * 10000)
```



### Task 5: Calculate the violent and property crime rates for each year (same unit)

```{r}
crimes_with_rates <- crimes |> mutate(violent_rate = violent_crime_total/population * 10000, property_rate = property_crime_total/population * 10000)
```



### Task 6: Describe our results

Write a few sentences about what you found in the data. What years had the highest crime rates? What years had the lowest crime rates? Does the pattern match your expectations?

1990 through 1996 had the highest violent crime rates, while the highest property crime rates were in 1980 and 1981. 2020 has both the lowest violent crime rate and property crime rate. Overall, it appears that the rates of both violent crime and property crime have been steadily decreasing since the late 90s. The high crime rates, particularly in the 90s were not very surprising to me. I know that certain socioeconomic factors and widespread issues like the crack epidemic led to an increase in youth and drug-related crime. However, I was not anticipating the relatively low rates in 2020. For some reason, the coverage of crime makes it feel like the problem has gotten worse. I have also heard some older people's opinions that crime is more of an issue now. I am not sure what that is based on. 


When you are done, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.

