---
title: "Hypothesis Testing & Effect Size"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Did Community Policing Actually Improve Response Times?

The city of Westbrook implemented a new community policing program in January. The police chief claims the program has improved emergency response times. Before the program, the average response time was **8.2 minutes** with a standard deviation of **1.5 minutes**. You've collected data from 20 randomly selected emergency calls after the program started.

Your editor wants to know two things: Is the improvement real? And is it big enough to matter?

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```

### Task 2: Set up the data and the hypotheses (2 points)

Before we run any tests, we need to state our hypotheses. The null hypothesis is what we assume to be true until we have evidence otherwise — like "innocent until proven guilty." Fill in the REPLACE_ME values.

**Null Hypothesis (H₀):** There is no statistically significant relationship between the new community policing program and emergency response times.

**Alternative Hypothesis (H₁):** With the implementation of the new community policing program, emergency response times have significantly improved/decreased.

```{r}
# Response times (in minutes) after the community policing program
response_times <- c(
  7.8, 8.1, 7.5, 7.9, 8.3,
  7.6, 7.4, 8.0, 7.7, 7.8,
  8.2, 7.3, 7.9, 7.5, 8.1,
  7.6, 7.8, 7.4, 7.7, 7.9
)

# Known baseline (before the program)
baseline_mean <- 8.2
baseline_sd <- 1.5

# Create a tibble
response_data <- tibble(
  call_number = 1:20,
  response_time = response_times
)

response_data
```

### Task 3: Calculate summary statistics (1 point)

Let's see what the new response times look like. Replace the REPLACE_ME values.

```{r}
summary_stats <- response_data |>
  summarize(
    sample_size = n(),
    sample_mean = mean(response_time),
    sample_sd = sd(response_time),
    standard_error = sample_sd / sqrt(sample_size)
  )

summary_stats
```

**Answer this question:** The sample mean is lower than the baseline of 8.2 minutes. How much lower? Does that seem like a meaningful improvement to you, just based on the numbers? What would you want to know before deciding?

The sample mean is a little roughly 0.4 minutes lower than the baseline mean, which is not a large decrease. However, the standard deviation for the sample size is much lower, demonstrating less variation in the data from the new response times. That is more meaningful than the decrease in the mean, in my opinion.

### Task 4: Perform a one-sample t-test (2 points)

Now we formally test the police chief's claim. Since we're testing whether response times *decreased*, we use a one-sided test (`alternative = "less"`). Replace the REPLACE_ME with the baseline mean we're comparing against.

```{r}
# Set our significance level
alpha <- 0.05

# Perform the t-test
t_test_result <- t.test(
  response_data$response_time,
  mu = 8.2,
  alternative = "less"
)

t_test_result
```

### Task 5: Interpret the t-test results (2 points)

Let's pull out the key numbers. Look at the p-value: is it less than our alpha of 0.05?

```{r}
# Extract key values
p_value <- t_test_result$p.value

results <- tibble(
  `Baseline mean` = paste(baseline_mean, "minutes"),
  `New mean` = paste(round(summary_stats$sample_mean, 2), "minutes"),
  `Difference` = paste(round(baseline_mean - summary_stats$sample_mean, 2), "minutes"),
  `p-value` = round(p_value, 6),
  `Significant?` = if_else(p_value < alpha, "Yes - reject null hypothesis", "No - fail to reject")
)

results
```

**Answer this question:** Based on the p-value, can we reject the null hypothesis? What does this tell the police chief — and what does it NOT tell him?

The p-value indicates that there is a significant difference between the baseline mean and the new mean, so we can reject the null hypothesis. This tells the police chief that there has been significant change in the response times, however, it does not tell how impactful the change is.

### Task 6: Visualize the results

```{r}
ggplot(response_data, aes(x = response_time)) +
  geom_histogram(binwidth = 0.2, fill = "steelblue", color = "white", alpha = 0.7) +
  geom_vline(xintercept = baseline_mean, color = "red", linetype = "dashed", size = 1.2) +
  geom_vline(xintercept = summary_stats$sample_mean, color = "darkgreen", size = 1.2) +
  annotate("text", x = baseline_mean + 0.05, y = 4,
           label = paste("Baseline:", baseline_mean),
           color = "red", hjust = 0, fontface = "bold") +
  annotate("text", x = summary_stats$sample_mean - 0.05, y = 3.5,
           label = paste("New Mean:", round(summary_stats$sample_mean, 2)),
           color = "darkgreen", hjust = 1, fontface = "bold") +
  labs(
    title = "Emergency Response Times: Before vs. After Community Policing",
    subtitle = paste0("p-value = ", round(p_value, 6)),
    x = "Response Time (minutes)",
    y = "Number of Calls"
  ) +
  theme_minimal()
```

### Task 7: Calculate the effect size (3 points)

The t-test told us the improvement is statistically significant — it's real, not due to chance. But is it *meaningful*? That's what effect size measures. We'll calculate **Cohen's d**, which tells us how large the difference is in standard deviation units.

The formula is: d = (baseline mean - sample mean) / baseline standard deviation

Replace the REPLACE_ME values.

```{r}
effect_size <- (8.2 - summary_stats$sample_mean) / 1.5

effect_size_interpretation <- case_when(
  abs(effect_size) < 0.2 ~ "Negligible",
  abs(effect_size) < 0.5 ~ "Small",
  abs(effect_size) < 0.8 ~ "Medium",
  TRUE ~ "Large"
)

tibble(
  `Cohen's d` = round(effect_size, 3),
  `Interpretation` = effect_size_interpretation,
  `Difference in minutes` = round(baseline_mean - summary_stats$sample_mean, 2),
  `Difference in seconds` = round((baseline_mean - summary_stats$sample_mean) * 60, 0)
)
```

**Answer this question:** The result is statistically significant, but the effect size is small. In practical terms, the improvement is about 25 seconds per call. The police chief is claiming success. Would you write a story saying the program works? How would you frame the findings honestly — acknowledging both that the improvement is real AND that it's small?

I do not know that I would write a story lauding the program as changing response times given that the actual decrease is very small. I think I would write a story about the implementation of the new program and note that response times decreased marginally, but I would not make the headline or the center of the story lower response times given that in reality, the difference is not impactful. I might note the difference in standard deviation, showing that response times have become somewhat more consistent.

### Task 8: The journalist's framework (5 points)

The videos you watched describe two questions you need to ask about any statistical result:

1. **Is it statistically significant?** (Is the result unlikely to be due to chance?)
2. **Is it meaningfully significant?** (Is the effect large enough to matter?)

Apply this framework to the community policing data:

**Question 1:** Write 1-2 sentences answering whether the result is statistically significant and how you know.

The results are statistically significant, because the p-value we received from the t-test was very small.

**Question 2:** Write 1-2 sentences answering whether the result is meaningfully significant. Consider: the program cost the city $2 million to implement. Is a 25-second improvement in response times worth that investment? What context would you need to answer that question?

The result, a 25 second decrease in average emergency response time, is not meaningfully significant, especially when the original time is over 8 minutes. Because the program costs $2 million, I do not think the results show that this program is worth the investment, or that it should be praised for decreasing response times. I suppose the program could be worth it if there are more advantages than just the decreased response times.

**Question 3:** Write a headline and a 2-3 sentence lede for a story about these findings. Your text should accurately reflect both the statistical significance and the effect size — don't let the police chief's framing drive the story if the data tells a more nuanced one.

Headline: Westbrook's New Community Policing Program Has a Marginal Impact on Emergency Response Times

Lede: A recent study of Westbrook's response data shows that emergency response times have decreased about 25 seconds since the implementation of the community policing program. The police chief has promoted this program by lauding the improved response times. However, while statistically significant, many wonder whether a 25 second improvement is worth a $2 million investment.


When you are finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
