---
title: "Identifying Schools with Biggest Differences in Math Test Scores, Part 2"
author: "Lauren Smith"
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
options(scipen = 999)
```

### Introduction

In the class activity, you tested a school board member's claim using data from Montgomery County, Pa. — and found a model that didn't hold up. Now you'll run the same analysis using data from Montgomery County, Md. Your job is to evaluate whether the Maryland data tells a different story, and what that means for how you'd report on the claim.

### Loading the Data

```{r}
schools <- read_csv("https://raw.githubusercontent.com/dwillis/jour405_files/refs/heads/main/montgomery_md.csv")
```

### Task 1: Examine the Data First (2 points)

Before you plot anything, look at the data — you know why by now.

```{r}
summary(schools$`FARMS Pct`)
schools |> arrange(desc(`FARMS Pct`)) |> head(10)
```

Are there any data quality issues like the ones you found in the Pennsylvania data? How does the distribution of FARMS percentages compare between the two counties? Note anything that concerns you or gives you confidence in the data.

I don't see any data quality issues like in the Pennsylvania data. Unlike the former dataset, there is no FARMS percentage of 100 and no outrageously large gap in the FARMS percentages at any point. The distribution of values appears much more normal throughout, which makes me feel more confident in the data than with the last county.


### Task 2: Create a Scatterplot and Compare (5 points)

Replace the REPLACE_ME value with the correct variable name to create a scatterplot of FARMS percentage vs. math scores for Montgomery County, Md.

```{r}
schools |>
  ggplot(aes(x = `FARMS Pct`, y = `% Met or Exceeded Math Expectations`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "FARMS Percentage", y = "% Met or Exceeded Math Expectations",
       title = "Math Scores vs. FARMS Percentage, Montgomery County MD")
```

Describe what you see. Then compare this scatterplot specifically to the Pennsylvania one: what differences do you notice in the direction, strength, and spread of the relationship? The two counties share a name and are both suburban — what might explain why they look so different?

This scatterplot shows a decrease in math scores as FARMS percentage increases. The data points are spread along the regression line going down to the right, indicating a strong negative relationship between the two variables. This is very different from the Pennsylvania scatterplot (without the 100% exclusion), where the regression line went up slightly, and the data points were not spread evenly along the line. The Pennsylvania regression line that included the suspicious data points implied a weak positive relationship, while the one that excluded the data points showed a moderate positive relationship. The second Pennsylvania scatterplot showed more evenly distributed data points.

One reason might be that the Maryland dataset has more values than the Pennsylvania dataset, which could mean more accurate results. The Maryland data set also does not have suspicious data points or gaps in the numbers that would skew the data. Also, outside of data reasons, there may be an overall difference in funding or resources between the two counties. If Montgomery County, Pa. has less students per class, there may be more time dedicated to helping each student. While there might be a correlation between FARMS percentage and math scores, even schools with more underprivileged students could produce better educational outcomes with the right support.

### Task 3: Linear Regression Model (5 points)

Replace the REPLACE_ME value with the correct variable name to run a linear regression model predicting math scores from FARMS percentage.

```{r}
model <- lm(`% Met or Exceeded Math Expectations` ~ `FARMS Pct`, data = schools)
summary(model)
```

Report the three key statistics: the p-value, the Adjusted R-squared, and the coefficient for FARMS Pct. Then interpret them together — not separately. What does the combination of these three numbers tell you about the relationship between poverty and math scores in Montgomery County, Md.? Is this model worth using? How does it compare to the Pennsylvania model?

The p-value is very low, indicating an undeniably statistically significant relationship between FARMS percentage and math scores. The r-squared at 0.6734 means that a good amount, roughly 67%, of the variance in math scores of Montgomery County, Md. elementary schools is explained by FARMS percentage. The coefficient, -0.59511, indicates a negative relationship between the two variables, meaning the higher the poverty, the lower the math scores in Montgomery County, Md. The coefficient is also farther away from 0, indicating a stronger relationship.

I feel like this model is worth using. It shows a much stronger and more significant relationship than the Pennsylvania models.


### Task 4: Identifying Schools with the Biggest Residuals (6 points)

Replace the REPLACE_ME values with the correct variable names to calculate predicted values, residuals, and absolute residuals for each school. Then arrange the schools in descending order of absolute residuals.

```{r}
schools <- schools |>
  mutate(
    predicted = predict(model),
    residual = `% Met or Exceeded Math Expectations` - predicted,
    abs_residual = abs(residual)
  ) |>
  arrange(desc(abs_residual))

head(schools |> select(1, `FARMS Pct`, `% Met or Exceeded Math Expectations`, predicted, residual), 10)
```

Identify the school with the largest positive residual (doing much better than the model predicts) and the school with the largest negative residual (doing much worse). For each school, find its zip code and look it up on [CensusReporter.org](https://censusreporter.org).

For each of the two schools, answer:
- What does the demographic and economic profile of the zip code look like?
- Does anything in that profile help explain why the school is over- or under-performing the model?
- What would your first reporting question be to the principal of that school?

The school with the largest positive residual, with 29.99, is Fox Chapel Elementary. The population of the zip code is 26,406 people, with about 9.8% of the population living below the poverty line. The demographics show Black and Hispanic populations with significant White and Asian populations. I do not see anything specifically in the demographics that would explain over-performance. My first question would be about the resources and funding at the school. I would want to see if there were extra programs for math or if teachers had more time to dedicate to the students

The school with the largest negative residual, with -28.08, is Germantown Elementary. The population of this zip code is 62,738 people with 8.3% of people living below the poverty line. The demographics show that the population is largely white and black, with significant Hispanic and Asian populations. I do not see anything glaring in the demographics that would explain under-performance. My first question to the principal would be about the student to teacher ratio, and overall resources. The population of this zip code is almost 3x the population of the zip code holding the over-performing school. This may mean more students to each teacher, which makes it hard for students who are having difficulty in math to receive the support they need.


### Task 5: Evaluate the Model and Write the Pitch (7 points)

You've now run this analysis on two counties. Return to the school board member's claim: that elementary schools with fewer students on free and reduced-price meals do better on standardized math tests.

Write a short story pitch of 4-6 sentences that addresses all of the following: Does the Maryland data support or contradict the claim? How does it compare to what you found in Pennsylvania? What would you ask the school board member if you called them for comment? And what's the most interesting specific finding from your residuals analysis: which school would you want to visit first, and why?

In contrast to the data from Montgomery County, Pa., the Montgomery County, Md. data supports the Pennsylvania school board member's claim that elementary school with fewer FARMS eligible students perform better on standardized math tests. I would like to inform the school board member with the original claim about my findings in the data and ask what information he used to support his claim. The residuals analysis showed two schools in Montgomery County, Md. with relatively similar demographics, but one over-performed and another under-performed. The only difference is one of the schools is in a zip code with a smaller population. I would want to visit Fox Chapel Elementary, which has a higher FARMS percentage, and yet is over-performing. There is clearly some difference in resources or teaching dynamics that is causing over-performance.
